---
title: "NG00020 phenotype QC & prep"
author: Jai Broome
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    code_folding: show

*pulled from /nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/pheno/pheno_qc.html on 12/14, edits by Hanley Kingston
---

# Setup

```{r setup, message = FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(magrittr)
library(survival)
library(knitr)
library(Biobase)
library(SeqArray)
library(blPipeline)
```

# Read data

```{r read_data}
sampled <- read.csv("/nfs/beluga0_home/DATA/NIAGADS/NG00020/NG00020_LOAD_phenotype_files/NG00020_LOAD_phenotype/phenotype/2009_11_16 Age and Source File.csv")

pheno <- read.csv("/nfs/beluga0_home/DATA/NIAGADS/NG00020/NG00020_LOAD_phenotype_files/NG00020_LOAD_phenotype/phenotype/2009_08_24_phenotype_data.csv")

dup_vars <- names(pheno)[names(pheno) %in% names(sampled)]
dup_vars <- dup_vars[dup_vars != "SUBJ_NO"]

apoe <- read.csv("/nfs/beluga0_home/DATA/NIAGADS/NG00020/NG00020_LOAD_phenotype_files/NG00020_LOAD_phenotype/phenotype/2009_12_17 APOE results.csv") %>%
  mutate(E2 = (Translated_Allele1 == "E2") + (Translated_Allele2 == "E2"),
         E3 = (Translated_Allele1 == "E3") + (Translated_Allele2 == "E3"),
         E4 = (Translated_Allele1 == "E4") + (Translated_Allele2 == "E4"))

dd <- file.path("/nfs/beluga0_home/DATA/NIAGADS/NG00020/NG00020_LOAD_phenotype_files/NG00020_LOAD_phenotype/phenotype/2009_07_24_data_dictionary.csv") %>%
  read.csv() %>%
  # Subset to just variables we're using for this ADF
  filter(VARNAME %in% c(names(pheno), names(sampled)), VARNAME != "SUBJ_NO") %>%
  select(VARNAME, labelDescription = VARDESC) %>%
  rbind(c(VARNAME = "AgeAtLastEval", labelDescription = "Age at last evaluation"))

dd_pheno <- dd_sampled <- filter(dd, VARNAME %in% dup_vars)

dd_pheno$VARNAME %<>% paste0("_pheno")
dd_pheno$labelDescription %<>% paste("(from 2009_08_24_phenotype_data.csv)")
dd_sampled$VARNAME %<>% paste0("_source")
dd_sampled$labelDescription %<>% paste("(from 2009_11_16 Age and Source File.csv)")

dd %<>% filter(!(VARNAME %in% dup_vars)) %>%
  bind_rows(dd_pheno, dd_sampled)# %>%
  #rbind(dd_pcs)


pheno %<>% full_join(sampled, "SUBJ_NO", suffix = c("_pheno", "_source")) %>%
  #full_join(pcs, "SUBJ_NO") %>%
  full_join(apoe, c(SUBJ_NO = "CIDR2_Subj_No")) %>%
  mutate(across(matches("age"), ~ifelse(.x == 999, NA, .x))) %>%
  # Several age variables are in both datasets. Create index where they don't match
  mutate(AgeDxDemMatch = AgeDxDem_pheno == AgeDxDem_source,
         AgeAtLastEvalMatch = AgeAtLastEval_pheno == AgeAtLastEval_source)

```

# Make age variable

Create a simple function that takes a variable name and makes that the age
variable for any missing observations in `x$age`, and records the source.

```{r make_age_var, warning = FALSE}
pheno$age <- NA_integer_
pheno$age_source <- NA_character_
make_age_var <- function(x, y){
  i <- is.na(x$age)
  x$age[i] <- x[[y]][i]
  x$age_source[i] <- y
  x
}
# Specify the order we want the age variable for cases
cases <- filter(pheno, Case_Control == "1") %>%
  make_age_var("AgeDem_pheno") %>%
  make_age_var("AgeDem_source") %>%
  make_age_var("AgeDxDem_source") %>%
  make_age_var("AgeDxDem_pheno") %>%
  make_age_var("Sampled.Age") %>%
  make_age_var("AgeAtLastEval_pheno") %>%
  make_age_var("AgeAtLastEval_source") %>%
  make_age_var("AgeDeath") %>%
  # Only one observation mismatched and it's already flagged, so min/max order
  # doesn't matter.
  mutate(dx_age_diff = pmin(AgeDxDem_pheno, AgeDxDem_source, na.rm = TRUE) -
                       pmax(AgeDem_pheno, AgeDem_source, na.rm = TRUE),
         # Indicator for when Dx is before symptoms
         dx_age_flag = dx_age_diff < 0,
         # Indicator for when lag between symptoms and Dx is greater than 10
         # years
         long_lag_dx_flag = dx_age_diff > 10,
         control_with_dem = NA)
# Specify order for controls
controls <- filter(pheno, Case_Control == "0") %>%
  make_age_var("AgeAtLastEval_source") %>%
  make_age_var("AgeAtLastEval_pheno") %>%
  make_age_var("Sampled.Age") %>%
  make_age_var("AgeDeath") %>%
  rowwise() %>%
  mutate(dx_age_diff = NA, dx_age_flag = NA, long_lag_dx_flag = NA,
         # Indicator for controls that have an age-at-sympmtom observation
         control_with_dem = !all(is.na(AgeDem_pheno), is.na(AgeDem_source),
                                 is.na(AgeDxDem_pheno), is.na(AgeDxDem_source)))
# Recombine cases and controls
rec <- rbind(cases, controls)
```

# Age variable QC

## Age source by case/control status

```{r age_source}
age_source.table <- select(rec, age_source, Case_Control) %>% table(useNA = "ifany")
write.table(age_source.table, "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/age_source_table.txt")

```

## Variable discrepency

There are a handful of cases where variables that appear in both datasets have
different values:

```{r age_scatter}
dat_plot <- select(rec, SUBJ_NO, ends_with("_pheno"), ends_with("_source"),
       -starts_with("BirthYr"), -age_source) %>%
  pivot_longer(!SUBJ_NO, c("age_var", ".value"), names_sep = "_") %>%
  filter(pheno != source, !is.na(pheno), !is.na(source)) %>%
  left_join(pheno, "SUBJ_NO")
my_lims <- c(min(dat_plot$pheno, dat_plot$source),
             max(dat_plot$pheno, dat_plot$source))

png("/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/age_discrepancy_between_files.png")
ggplot(dat_plot, aes(x = pheno, y = source, color = as.factor(Case_Control))) +
  geom_point() +
  facet_grid(cols = vars(age_var)) +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  coord_fixed() +
  xlim(my_lims) +
  ylim(my_lims)
dev.off()
```

`AgeAtLastEval` was the age variable used for all 15 instances where it didn't
match between the two files

```{r age_test_qc}
rec$pass_qc <- TRUE
select(rec, age_source, AgeAtLastEvalMatch) %>%
  filter(!AgeAtLastEvalMatch) %>%
  kable()
ind <- !rec$AgeAtLastEvalMatch & rec$age_source == "AgeAtLastEval_source"
ind[is.na(ind)] <- FALSE
rec$pass_qc[ind] <- FALSE
```

## Age-at-symptoms and age-at-Dx

There are 12 observations with diagnosis _before_ onset of symptoms

```{r dx_before_symptoms}
select(cases, AgeDxDem_pheno, AgeDxDemMatch, AgeDem_pheno, dx_age_diff,
       dx_age_flag) %>%
  filter(dx_age_diff < 0) %>%
  kable()
```

These are flagged here along with those where there was a ten year gap between onset
of symptoms and a diagnosis. The grey line is the identity line and the black
line marks a difference of ten years.

```{r plot_dx_age_diff, warning = FALSE}
png("/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/age_DxvsAge_Dem.png")
mutate(cases, flag = dx_age_flag | long_lag_dx_flag) %>%
  ggplot(aes(x = AgeDem_pheno, y = AgeDxDem_pheno, color = flag)) +
  geom_jitter() +
  geom_abline(slope = 1, intercept = 0, color = "grey") +
  geom_abline(slope = 1, intercept = 10, color = "black") +
  coord_fixed()
dev.off()
```

And here is the difference for all subjects, not just those flagged:

```{r hist_diff}
png("/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/age_DxvsAge_Dem_hist.png")
ggplot(cases, aes(x = dx_age_diff)) + geom_histogram(binwidth = 1) +
  geom_vline(xintercept = 0, color = "red")
dev.off()
```

Note the steep dropoff at exactly the 10 year mark.

```{r dx_age_flag}
ind <- rec$dx_age_flag
ind[is.na(ind)] <- FALSE
rec$pass_qc[ind] <- FALSE
```

```{r long_lag_dx_flag}
ind <- rec$long_lag_dx_flag
ind[is.na(ind)] <- FALSE
rec$pass_qc[ind] <- FALSE
```

## 3 controls have a age-at-dementia or age-at-Dx variable

```{r}
filter(controls, control_with_dem) %>%
  select(matches("agedem"), matches("agedx"), control_with_dem) %>%
  kable()
ind <- filter(controls, control_with_dem)$SUBJ_NO
rec$pass_qc[rec$SUBJ_NO %in% ind] <- FALSE
```

## The mismatched `AgeDem` variable wasn't used:

```{r}
select(rec, age_source, AgeDxDemMatch) %>%
  filter(!AgeDxDemMatch) %>%
  kable()
```

```{r}
rec[!rec$pass_qc, ] %>%
  group_by(Case_Control, age_source, dx_age_flag, long_lag_dx_flag,
           control_with_dem, AgeDxDemMatch, AgeAtLastEvalMatch) %>%
  summarize(n = n()) %>%
  kable()

#Can't figure out how to save this, so copied and pasted it into thi file:
#save_kable(source_of_flag, "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/pheno_flagged_samples.txt")
```

# APOE allele QC

Confirm that E2/E3/E4 allele counts all make sense

```{r apoe_qc}
group_by(rec, Translated_Allele1, Translated_Allele2, E2, E3, E4) %>%
  summarize(n = n())

allele_counts <- group_by(rec, Translated_Allele1, Translated_Allele2, E2, E3, E4) %>%
  summarize(n = n())

allele_counts

write.table(allele_counts, "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/allele_coutns_table.txt")
```


# Survival

```{r}
gds <- seqOpen("/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/data/D_LOAD_families_010413_aut.gds")
gds.id <- seqGetData(gds, "sample.id")
ids.missing <- gds.id[!(gds.id %in% rec$SUBJ_NO)]
seqClose(gds)
rec %<>% bind_rows(data.frame(SUBJ_NO = ids.missing))
covars <- c("E2", "E4", "E3")
rec <- rec[match(gds.id, rec$SUBJ_NO), ] %>%
  make_resids("age", "Case_Control", covars = c("SEX", covars))
m <- filter(rec, SEX == 1) %>%
  make_resids("age", "Case_Control", "surv_sex_strat", covars,
              mgl_name = "mgl_sex_strat", dev_name = "dvr_sex_strat")
rec %<>% filter(SEX == 2 | is.na(SEX)) %>%
  make_resids("age", "Case_Control", "surv_sex_strat", covars,
              mgl_name = "mgl_sex_strat", dev_name = "dvr_sex_strat") %>%
  rbind(m)
```

# Write out AnnotatedDataFrame

## Change sample ID variable name to expected value

```{r sampname}
rec %<>% rename(sample.id = SUBJ_NO)
```

## reformat sex, race, and ethnicity variables for plotting
rec <- rec %>%
    mutate(SEX=c("1"="M", "2"="F")[as.character(SEX)],
           Race=c("1"="White", "2"="Black", "3"="AI_AN", "4"="Asian_PI", "50"="Other")[as.character(Race)],
           Hispanic=c("1"="yes", "2"="no")[as.character(Hispanic)],
			  race_ethnicity = case_when(
					(Race == "Other" | Race == "AI_AN" | Race == "Asian_PI") & Hispanic == "yes" ~ "Other_Hispanic",
					Race == "Black" & Hispanic == "yes" ~ "Black_Hispanic",
					Race == "White" & Hispanic == "yes" ~ "White_Hispanic",
					TRUE ~ Race
           )
		)

## Make E2 and E4 carrier and het/hom columns
rec <- rec %>%
	mutate(
		E2_carrier = case_when(E2 == 1 | E2 == 2 ~ "yes",
									  E2 == 0 ~ "no"),
		E4_carrier = case_when(E4 == 1 | E4 == 2 ~ "yes",
									  E4 == 0 ~ "no"),
		E2_het_hom = case_when(E2 == 1 ~ "heterozygote",
									  E2 == 2 ~ "homozygote"),
		E4_het_hom = case_when(E4 == 1 ~ "heterozygote",
									  E4 == 2 ~ "homozygote")
			)

rec[!rec$pass_qc, ] %>%
  group_by(E2, E4, E2_carrier, E4_carrier, E2_het_hom, E4_het_hom) %>%
  summarize(n = n()) %>%
  kable()

#saved to: "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/logical_APOE_genotype_check.txt"


```{r}
dd %<>% rbind(
    c("sample.id", "Sample ID"),
    # Some variables we started with aren't defined in the DD. I haven't
    # attempted to define them here.
	 c("race_ethnicity", ""),
	 c("E2_carrier", ""),
	 c("E4_carrier", ""),
	 c("E2_het_hom", "non-carriers are NA"),
	 c("E4_het_hom", "non-carriers are NA"),
    c("EvalYr", ""),
    c("EvalMeth", ""),
    c("Sampled.Age", ""),
    c("Sample.Type", ""),
    c("AgeDxDemMatch", "AgeDxDem is the same in both datasets"),
    c("AgeAtLastEvalMatch", "AgeAtLastEval is the same in both datasets"),
    c("age", "Selected age for cases and controls. See phenotype QC for details."),
    c("age_source", "Source variable and dataset for variable `age`"),
    c("dx_age_diff", "Difference between age at AD Dx and dementia symptoms"),
    c("dx_age_flag", "Flag for cases where onset of symptoms is recorded as after diagnosis"),
    c("long_lag_dx_flag", "Flag for cases with greater than ten year lag between onset of dementia symptoms and AD Dx"),
    c("control_with_dem", "Controls with some recorded 'age-at-dem' variable"),
    c("pass_qc", "Passes all flags"),
    c("surv", "output from survival::Surv()"),
    c("devres", "Deviance residuals from survival::residuals()"),
    c("martingale", "Martingale residuals from survival::residuals()"),
    c("surv_sex_strat", "sex-stratified output from survival::Surv()"),
    c("dvr_sex_strat", "sex-stratified deviance residuals"),
    c("mgl_sex_strat", "sex-stratified martingale residuals"),
    c("rs429358.Allele_1", ""),
    c("rs429358.Allele_2", ""),
    c("Confidence", ""),
    c("rs7412.Allele1", ""),
    c("rs7412.Allele2", ""),
    c("Confidence2", ""),
    c("Translated_Allele1", ""),
    c("Translated_Allele2", ""),
    c("E2", "Count of APOE2 alleles"),
    c("E3", "Count of APOE3 alleles"),
    c("E4", "Count of APOE4 alleles")
  ) %>%
  as.data.frame()

if (length(names(rec)[!(names(rec) %in% dd$VARNAME)]) != 0) {
  stop("There are undefined variables")
}
if (length(dd$VARNAME[!(dd$VARNAME %in% names(rec))]) != 0) {
  stop("There are entries in the DD that are not in the dataset")
}

id.keep <- rec$sample.id[rec$pass_qc & !is.na(rec$pass_qc)]
length(id.keep)
#[1] 4412

saveRDS(id.keep, "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/pass_pheno_qc.rds")

vmd <- select(dd, labelDescription)
rownames(vmd) <- dd$VARNAME
pheno_adf <- AnnotatedDataFrame()
pData(pheno_adf) <- as.data.frame(rec)
varMetadata(pheno_adf) <- vmd

saveRDS(pheno_adf, "/nfs/beluga0_home/ANALYSIS/NIAGADS/NG00020/hkings/pheno/out/pheno_adf.rds")
```
